---
title: "2.7: Google Culture?"
author: "Alex Kindel"
date: "November 8, 2016"
output: html_document
---

```{r setup, include=FALSE}
require(dplyr)
require(ggplot2)
```

## Works referenced

Michel, Jean-Baptiste, Yuan Kui Shen, Aviva P. Aiden, Adrian Veres, Matthew K. Gray, The Google Books Team, Joseph P. Pickett, et al. 2011. “Quantitative Analysis of Culture Using Millions of Digitized Books.” Science 331 (6014): 176–82. doi:10.1126/science.1199644.

Pechenick, Eitan Adam, Christopher M. Danforth, and Peter Sheridan Dodds. 2015. “Characterizing the Google Books Corpus: Strong Limits to Inferences of Socio-Cultural and Linguistic Evolution.” PLoS ONE 10 (10): e0137041. doi:10.1371/journal.pone.0137041.

## Background

Michel et al. (2011) constructed a corpus emerging from Google’s effort to digitize books. Using the first version of the corpus, which was published in 2009 and contained over 5 million digitized books, the authors analyzed word usage frequency to investigate linguistic changes and cultural trends. Soon the Google Books Corpus became a popular data source for researchers, and a 2nd version of the database was released in 2012.

However, Pechenick, Danforth, and Dodds (2015) warned that researchers need to fully characterize the sampling process of the corpus before using it for drawing broad conclusions. The main issue is that the corpus is library-like, containing one of each book. As a result, an individual, prolific author is able to noticeably insert new phrases into the Google Books lexicon. Moreover, scientific texts constitute an increasingly substantive portion of the corpus throughout the 1900s. In addition, by comparing two versions of the English Fiction datasets, Pechenick et al. found evidence that insufficient filtering was used in producing the first version. All of the data needed for activity is available here: http://storage.googleapis.com/books/ngrams/books/datasetsv2.html

```{r fetch_data}
setwd("~/Code/css_activities")
system("./process_ngrams.py")
```

## A

In Michel et al.’s original paper (2011), they used the 1st version of the English data set, plotted the frequency of usage of the years “1880”, “1912” and “1973”, and concluded that “we are forgetting our past faster with each passing year” (Fig. 3A, Michel et al.). Replicate the same plot using 1) 1st version of the corpus, English dataset (same as Fig. 3A, Michel et al.)

## B

Now replicate the same plot with the 1st version, English fiction dataset.

## C

Now replicate the same plot with the 2nd version of the corpus, English dataset.

## D
Finally, replicate the same plot with the 2nd version, English fiction dataset.

## Discussion

Describe the differences and similarities between these four plots. Do you agree with Michel et al.’s original interpretation of the observed trend? (Hint: c) and d) should be the same as Figure 16 in Pechenick et al.)
Now that you have replicated this one finding using different Google Books corpora, choose another linguistic change or cultural phenomena presented in Michel et al.’s original paper. Do you agree with their interpretation in light of the limitations presented in Pechenick et al.? To make your argument stronger, try replicate the same graph using different versions of data set as above.